Analysis-ready VCF at Biobank scale using Zarr

### 2025-01-31 response-to-reviewers.tex

\begin{point}
The authors present VCF Zarr, a specification that translates the variant call
format (VCF) data model into an array-based representation for the Zarr storage
format. They also present the `vcf2zarr` utility to convert large VCFs to Zarr.
They provide data compression and analysis benchmarks comparing VCF Zarr to
existing variant storage technologies using simulated genotype data. They also
present a case study on real world Genomics England aggV2 data.

The authors' benchmarks overall show that VCF Zarr has superior compression and
computational analysis performance at scale relative to data stored as
row-oriented VCF and that VCF Zarr is competitive with specialized storage
solutions that require similarly specialized tools and access libraries for
querying. An attractive feature is that VCF Zarr allows for variant annotation
workflows that do not require full dataset copy and conversion. Another key
point is that Zarr is a high-level spec and data model for the chunked storage
of n-d arrays, rather than a byte-level encoding designed specifically around
the genomic variant data type. I personally have used Zarr productively for
several applications unrelated to statistical genetics. While Zarr VCF mildly
underperforms some of the specialized formats (Savvy in compute, Genozip in
compression) in a few instances, I believe the accessibility, interoperability,
and reusability gains of Zarr make the small tradeoff well worthwhile.

Because Zarr has seen heavy adoption in other scientific communities like the
geospatial and Earth sciences, and is well integrated in the scientific Python
stack, I think it holds potential for greater reusability across the ecosystem.
As such, I think the VCF Zarr spec is a highly valuable if not overdue
contribution to an entrenched field that has recently been confronted by a
scalability wall.

Overall, the paper is clear, comprehensive, and well written. 
\end{point}
\begin{reply}
Thank you!  We are delighted that you agree with us on the transformative
potential of VCF Zarr, and thank you for your insightful points and
helpful suggestions.
\end{reply}

\begin{point}
The benefits for large scientific datasets to be analysis-ready
cloud-optimized (ARCO) have been well articulated by Abernathey et al., 2021.
However, I do think that the "local"/HPC single-file use case is still
important and won't disappear any time soon, and for some file system use
cases, expansive and deep hierarchies can be performance limiting (this was
hinted at in one of the benchmarks). In this scenario would a large Zarr VCF
perform reasonably well (or even better on some file systems) via a single
local zip store?
\end{point}
\begin{reply}
This is good point, and we have added several mentions of the Zip file backend
as well as benchmarks using it. We added a new paragraph on pg5, (l318-338)
to discuss this point. We also added benchmarks based on the SARS-CoV-2
dataset, which is distributed as a Zipfile.
\end{reply}

\begin{point}
The description of the intermediate columnar format (ICF) used by `vcf2zarr`
is missing some detail. At first I got the impression it might be based on
something like Parquet, but running the provided code showed that it consists
of a similar file-based chunk layout to Zarr. This should be clarified in the
manuscript. 
\end{point}
\begin{reply}
We have tried to clarify this (and the purpose of ICF) by saying
``...storing each field independently in (approximately) fixed-size
compressed chunks in a file-system hierarchy.
ICF is designed to support efficient Zarr encoding within \texttt{vcf2zarr}
and not intended for reuse outside that context.''
\end{reply}

\begin{point}
The authors discuss the possibility of storing an index mapping genomic
coordinates to chunk indexes. Have Zarr-based formats in other fields like
geospatial introduced their own indexing approaches to take inspiration from?
\end{point}
\begin{reply}
We have added a new \texttt{region\_index} field for this purpose, which 
supports efficient genomic range searches. This is discussed in the last 
paragraph of the vcf2zarr Methods section (pg 13).
\end{reply}

\begin{point}
Since VCF Zarr is still a draft proposal, it could be useful to indicate
where community discussions are happening and how potential new contributors
can get involved, if possible. This doesn't need to be in the paper per se, but
perhaps documented in the spec repo.
\end{point}
\begin{reply}
Thank you, that is an excellent idea. We have added basic guidance for 
contributors to the specification repo.
\end{reply}

\begin{point}
In the background: "For the representation to be FAIR, it must also be
accessible," -- A is for "accessible", so "also" doesn't make sense.
\end{point}
\begin{reply}
Fixed
\end{reply}

\begin{point}
"There is currently no efficient, FAIR representation...". Just a nit and
feel free to ignore, but the solution you present is technically "current".
\end{point}
\begin{reply}
You are right, but for narrative purposes it is simpler to pose VCF Zarr 
as new, so we'd prefer to keep this as it is.
\end{reply}

\begin{point}
In Figure 2, the zarr line is occluded by the sav line and hard to see.
\end{point}
\begin{reply}
The benchmarks have changed slightly, and the lines are now more separated.
\end{reply}

### 2025-01-31 paper.tex
Reviewer 1 responses

To achieve this full potential we
need a new generation of tools that operate directly
on a primary data representation that supports
efficient access across a range of applications,
with native support for cloud object storage.
Such a representation can be termed ``analysis-ready''
and ``cloud-native''~\citep{abernathey2021cloud}.
For the representation to be FAIR~\citep{wilkinson2016fair},
it must also be \emph{accessible}, using protocols that are
``open, free, and universally implementable''.
There is currently no efficient, FAIR representation of genetic variation
data suitable for cloud deployments.
Hail~\cite{ganna2016ultra,hail2024} has become the dominant platform
for quality control of large-scale variation datasets,
and has been instrumental in projects such as
gnomadAD~\cite{karczewski2020mutational,chen2024genomic}.
While Hail is built on open components
from the Hadoop distributed computing ecosystem~\citep{white2012hadoop},
the details of its MatrixTable format are not documented
or intended for external reuse.
% cite? https://dev.hail.is/t/matrixtable-file-format-reference/173/6
Similarly, commercial solutions that have emerged to facilitate
the analysis of large-scale genetic variation data are either
based on proprietary~\cite{basespace2024,graf2024,googlelifesciences2024,
awshealthomics2024,microsoftgenomics2024}
or single-vendor technologies~\cite[e.g.][]{tiledb2024,genomicsdb2024}.
The next generation of VCF analysis methods requires
an open, free and transparent data representation
with multiple independent implementations.

@@

The \texttt{vcf2zarr} utility solves this problem by first converting
the VCF data (which can be split across many files) into an Intermediate
Columnar Format (ICF). The \texttt{vcf2zarr explode} command takes a set
of VCFs, and reads through them using
\texttt{cyvcf2}~\cite{pedersen2017cyvcf2},
storing each field independently in (approximately) fixed-size
compressed chunks.
Large files can be partitioned based on information extracted from the
CSI or Tabix indexes, and so different parts of a file can be
converted to ICF in parallel.
Once all partitions have completed, information about
the number of records in each partition and chunk of a given
field is stored so that the record at a particular index
can be efficiently retrieved.
Summaries such as maximum dimension and the minimum and maximum value
of each field are also maintained, to aid choice of data types later.
A set of VCF files can be converted to intermediate columnar
format in parallel on a single machine
using the \texttt{explode} command,
or can be distributed across a cluster using the
\texttt{dexplode-init},
\texttt{dexplode-partition} and \texttt{dexplode-finalise} commands.

>>>>

To achieve this full potential we
need a new generation of tools that operate directly
on a primary data representation that supports
efficient access across a range of applications,
with native support for cloud object storage.
Such a representation can be termed ``analysis-ready''
and ``cloud-native''~\citep{abernathey2021cloud}.
For the representation to be FAIR~\citep{wilkinson2016fair},
it must be \emph{accessible}, using protocols that are
``open, free, and universally implementable''.
There is currently no efficient, FAIR representation of genetic variation
data suitable for cloud deployments.
Hail~\cite{ganna2016ultra,hail2024} has become the dominant platform
for quality control of large-scale variation datasets,
and has been instrumental in projects such as
gnomadAD~\cite{karczewski2020mutational,chen2024genomic}.
While Hail is built on open components
from the Hadoop distributed computing ecosystem~\citep{white2012hadoop},
the details of its MatrixTable format are not documented
or intended for external reuse.
% cite? https://dev.hail.is/t/matrixtable-file-format-reference/173/6
Similarly, commercial solutions that have emerged to facilitate
the analysis of large-scale genetic variation data are either
based on proprietary~\cite{basespace2024,graf2024,googlelifesciences2024,
awshealthomics2024,microsoftgenomics2024}
or single-vendor technologies~\cite[e.g.][]{tiledb2024,genomicsdb2024}.
The next generation of VCF analysis methods requires
an open, free and transparent data representation
with multiple independent implementations.

@@

The \texttt{vcf2zarr} utility solves this problem by first converting
the VCF data (which can be split across many files) into an Intermediate
Columnar Format (ICF). The \texttt{vcf2zarr explode} command takes a set
of VCFs, and reads through them using
\texttt{cyvcf2}~\cite{pedersen2017cyvcf2},
storing each field independently in (approximately) fixed-size
compressed chunks in a file-system hierarchy.
ICF is designed to support efficient Zarr encoding within \texttt{vcf2zarr}
and not intended for reuse outside that context.
Large files can be partitioned based on information extracted from the
CSI or Tabix indexes, and so different parts of a file can be
converted to ICF in parallel.
Once all partitions have completed, information about
the number of records in each partition and chunk of a given
field is stored so that the record at a particular index
can be efficiently retrieved.
Summaries such as maximum dimension and the minimum and maximum value
of each field are also maintained, to aid choice of data types later.
A set of VCF files can be converted to intermediate columnar
format in parallel on a single machine
using the \texttt{explode} command,
or can be distributed across a cluster using the
\texttt{dexplode-init},
\texttt{dexplode-partition} and \texttt{dexplode-finalise} commands.
